---
title: "How to Conduct a Clone Censor-Weight Survival Analysis using survivalCCW"
author: Matthew Secrest
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Conduct a Clone Censor-Weight Survival Analysis using survivalCCW}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Clone Censor Weighting

This lightweight package describes how to conduct clone-censor weighting (CCW) to address the problem of immortal time bias in survival analysis. This vignette will walk through the applied tutorial published by [Maringe et al 2020](https://academic.oup.com/ije/article/49/5/1719/5835351). Refer to [Gaber et al 2024](https://doi.org/10.1007/s40471-024-00346-2) for more details on CCW in practice and to [Hernan and Robins 2016](https://doi.org/10.1093/aje/kwv254) for more technical details.

# Context

CCW is useful in the presence of immortal person-time bias in observational studies. For instance, when comparing surgery recipients vs non-recipients in non-small cell lung cancer (NSCLC), the surgery group will have a longer survival time than the non-surgery group because the non-surgery group includes patients who died before they could receive surgery. This is a form of immortal time bias.

The CCW toy dataset published by Maringe uses this exact setting as the motivating example. Let's explore the dataset, which comes with `survivalCCW`.

```{r, warning = FALSE}
library(survivalCCW)
library(DT) # For better printing of data.frames

head(toy_df) |>
  DT::datatable(
    rownames = FALSE, 
    options = list(
      scrollX = TRUE,
      paging = FALSE
    )
  )
```

Column descriptions can be found with `?toy_df`:

- `id`: patient identifier
- `fup_obs`: observed follow-up time (time to death or 1 year if censored alive)
- `death`: observed event of interest (all-cause death) 1: dead, 0: alive
- `timetosurgery`: time to surgery (NA if no surgery)
- `age`: age at diagnosis
- `sex`: patient's sex
- `perf`: performance status at diagnosis
- `stage`: stage at diagnosis
- `deprivation`: deprivation score
- `charlson`: Charlson's comorbidity index
- `emergency`: route to diagnosis

Note that this package addresses situations in which the covariates are all defined at baseline.

# Create clones

The first step is to create the clones. This can be done for any time-to-event outcome using the `survivalCCW` function `create_clones`. For `create_clones` to work, we need to pass a one-row-per-patient `data.frame` with the following columns:

- The id variable (in this case, `id`)
- The traditional outcome variable which denotes censorship (0) or event (1) (in this case, `death`). Note that additional values are not yet permitted.
- The time to the first event (in this case, `fup_obs`)
- The exposure variable, with exposure defined **at any time prior to censorship/event** (in this case, `surgery`). Must be (0) or (1),
- The time to exposure variable (in this case, `timetosurgery`)
- The clinical eligibility window (in this case, we'll do 6 months)

All other columns will be propogated for each patient. Let's see what this looks like in practice.

```{r}

# Create clones
clones <- create_clones(
  df = toy_df,
  id = 'id',
  event = 'death',
  time_to_event = 'fup_obs',
  exposure = 'surgery',
  time_to_exposure = 'timetosurgery',
  ced_window = 182.62
)

head(clones) |>
  DT::datatable(
    rownames = FALSE, 
    options = list(
      scrollX = TRUE,
      paging = FALSE
    )
  )
```


Note that this object is just a `data.frame` with an additional custom class which future functions will evaluate:
```{r}
class(clones)
```

You can visualize the censoring over time after you create the clones:

```{r}
plot_censoring_over_time(clones)
```


# Cast to long format
Now we simply need to cast the data to long format. The `survivalCCW` function `cast_to_long` will do this for us. 
No additional arguments are needed (the `clones` object is an artifact that allows you to better see and understand the method):

```{r}
clones_long <- cast_clones_to_long(clones)

head(clones_long, row.names = FALSE) |>
  DT::datatable(
      rownames = FALSE, 
      options = list(
        scrollX = TRUE,
        paging = FALSE
      )
    )
```

Let's pick out a single patient and look at their data:

```{r}
clones_long[clones_long$id == "P5913", ] |>
  DT::datatable(
    rownames = FALSE, 
    options = list(
      scrollX = TRUE,
      paging = TRUE
    )
  )
```

# Generate weights

Now we simply need to generate the weights. The `survivalCCW` function `generate_ccw()` will do this for us.

```{r}
clones_long_weights <- generate_ccw(clones_long, predvars = c("age", "sex", "perf", "stage", "deprivation", "charlson", "emergency"))

head(clones_long_weights) |>
  DT::datatable(
    rownames = FALSE, 
    options = list(
      scrollX = TRUE,
      paging = FALSE
    )
  )
```

Let's pick out a single patient and look at their data:

```{r}
clones_long_weights[clones_long_weights$id == "P5913", ] |>
  DT::datatable(
    rownames = FALSE, 
    options = list(
      scrollX = TRUE,
      paging = TRUE
    )
  )
```

You can also visualize weights over time with `plot_ccw_over_time()` and mean values over time with
`plot_var_mean_over_time()`.

```{r}
clones_long_weights |>
  plot_ccw_over_time()
```

```{r}
clones_long_weights |>
  plot_var_mean_over_time("age")
```

# Evaluate outcomes 

We now have everything we need to conduct a CCW analysis. For instance, we can pipe things together to evaluate the hazard ratio for surgery vs no surgery:

```{r}
library(survival)
df <- toy_df |>
  create_clones(
    id = 'id',
    event = 'death',
    time_to_event = 'fup_obs',
    exposure = 'surgery',
    time_to_exposure = 'timetosurgery',
    ced_window = 365.25/2
  ) |>
  cast_clones_to_long() |>
  generate_ccw(c('age', 'sex', 'perf', 'stage', 'deprivation', 'charlson', 'emergency'))

coxph(Surv(t_start, t_stop, outcome) ~ clone, data = df, weights = weight_cox)
```

Note that we used `outcome` and not `death` in the `coxph()` model. Still, there is of course a problem with this analysis, as the cloning process renders the variance invalid. The simplest approach to addressing this is 
to bootstrap the variance. I have not made a function to do this yet, but leave the below as an example of how to do this.

```{r}
library(boot)

boot_cox <- function(data, indices) {
  
  # Make long data.frame with weights
  ccw_df <- data[indices, ] |>
    create_clones(
      id = 'id',
      event = 'death',
      time_to_event = 'fup_obs',
      exposure = 'surgery',
      time_to_exposure = 'timetosurgery',
      ced_window = 182.62
    ) |>
    cast_clones_to_long() |>
    generate_ccw(c('age', 'sex', 'perf', 'stage', 'deprivation', 'charlson', 'emergency'))
    

  # Extract HR from CoxPH
  cox_ccw <- coxph(Surv(t_start, t_stop, outcome) ~ clone, data = ccw_df, weights = weight_cox) 
   
  hr <- cox_ccw |>
    coef() |>
    exp()

  out <- c("hr" = hr)

  # Create survfit objects for each of treated and untreated
  surv_1 <- survfit(Surv(t_start, t_stop, outcome) ~ 1L, data = ccw_df[ccw_df$clone == 1, ], weights = weight_cox)

  surv_0 <- survfit(Surv(t_start, t_stop, outcome) ~ 1L, data = ccw_df[ccw_df$clone == 0, ], weights = weight_cox)

  # RMST difference
  rmst_1 <- surv_1 |>
    summary(rmean = 365) |>
    (\(summary) summary$table)() |>
    (\(table) table["rmean"])()
  
  rmst_0 <- surv_0 |>
    summary(rmean = 365) |>
    (\(summary) summary$table)() |>
    (\(table) table["rmean"])()

  rmst_diff <- rmst_1 - rmst_0

  out <- c(out, "rmst_diff" = rmst_diff)

  # 1-year survival difference
  # Find the index of the time point closest to 1 year
  index_1yr_1 <- which.min(abs(surv_1$time - 365))
  index_1yr_0 <- which.min(abs(surv_0$time - 365))

  # Get the 1-year survival probabilities
  surv_1_1yr <- surv_1$surv[index_1yr_1]
  surv_0_1yr <- surv_0$surv[index_1yr_0]

  surv_diff_1yr <- surv_1_1yr - surv_0_1yr

  out <- c(out, "surv_diff_1yr" = surv_diff_1yr)

}

boot_out <- boot(data = toy_df, statistic = boot_cox, R = 10)
```

## Hazard ratios
```{r}
boot.ci(boot_out, type = "norm", index = 1)
```

## RMST
```{r}
boot.ci(boot_out, type = "norm", index = 2)
```

## 1-year survival
```{r}
boot.ci(boot_out, type = "norm", index = 3)
```


# Competing risks
The extension of the package functionality to issues of competing risks is trivial. We keep our outcome variable 
as an integer but add additional values.

```{r}
toy_df$competing <- sample(0:2, nrow(toy_df), replace = TRUE)
head(toy_df) |>
  DT::datatable(
    rownames = FALSE, 
    options = list(
      scrollX = TRUE,
      paging = FALSE
    )
  )
```

We can then conduct the same analysis as before, but with the `competing` variable as the outcome. The 
bootstrapped analysis is not shown, but you can see that hte same approach can be used.

```{r}
compete_ccw <- toy_df |>
  create_clones(
    id = 'id',
    event = 'competing',
    time_to_event = 'fup_obs',
    exposure = 'surgery',
    time_to_exposure = 'timetosurgery',
    ced_window = 365.25/2
  ) |>
  cast_clones_to_long() |>
  generate_ccw(c('age', 'sex', 'perf', 'stage', 'deprivation', 'charlson', 'emergency'))

table(compete_ccw$outcome)

head(compete_ccw) |>
  DT::datatable(
    rownames = FALSE, 
    options = list(
      scrollX = TRUE,
      paging = FALSE
    )
  )
```


# Winsorizing weights

You can trim/winsorize extreme weights using the `winsorize_ccw_weights()` function:

```{r}
compete_ccw_trim <- compete_ccw |>
  winsorize_ccw_weights(quantiles = c(0.10, 0.90))

max(compete_ccw$weight_cox)
max(compete_ccw_trim$weight_cox)
```

# References

Hernán, Miguel A., and James M. Robins. "Using big data to emulate a target trial when a randomized trial is not available." American journal of epidemiology 183.8 (2016): 758-764.

Gaber, Charles E., et al. "The Clone-Censor-Weight Method in Pharmacoepidemiologic Research: Foundations and Methodological Implementation." Current Epidemiology Reports (2024): 1-11.

Maringe, Camille, et al. "Reflection on modern methods: trial emulation in the presence of immortal-time bias. Assessing the benefit of major surgery for elderly lung cancer patients using observational data." International journal of epidemiology 49.5 (2020): 1719-1729.
